---
title: "Data Tidying"
author: "Jeanette Clark"
date: "10/16/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning = F, message = F}
library(dplyr)
library(tidyr)
library(DT)
```

Read in catch data from [Mike Byerly. Alaska commercial salmon catches by management region (1886- 1997). Gulf of Alaska Data Portal. df35b.304.2.](https://knb.ecoinformatics.org/view/df35b.304.2)

```{r}
catch_df <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"),
                    stringsAsFactors = FALSE)
head(catch_df)
```

`select` the columns we need for analysis
`gather` to move from wide to long
`rename` a column to indicate units

```{r}
catch_clean <- catch_df %>% 
  #select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
  select(-All, -notesRegCode) %>% 
  gather(key = "species", value = "catch", -Region, -Year) %>% 
  rename(catch_thousands = catch)

tail(catch_clean)
```

```{r}
test <- as.numeric(catch_clean$catch_thousands)
i <- which(is.na(test) == T)
catch_clean[i, ]
```

fix typeo in original dataset
coerce to catch to numeric, convert to number of fish
`mutate` to create a new catch column

```{r}
catch_clean_mutated <- catch_clean %>%
  # change "I" to 1 in catch_thousands column
  mutate(catch_thousands = ifelse(catch_thousands == "I", 1, catch_thousands)) %>%
  mutate(catch_thousands = as.numeric(catch_thousands)) %>% 
  mutate(catch = catch_thousands * 1000) %>% 
  select(-catch_thousands)

tail(catch_clean_mutated)
```

split using `group_by`
apply and combine using `summarize`

```{r}
catch_summarized <- catch_clean_mutated %>% 
  group_by(Region, Year) %>% 
  summarize(mean_catch = max(catch),
            n_obs = n(),
            stdev_catch = sd(catch))

head(catch_summarized)
```

```{r}
catch_chinook <- catch_clean_mutated %>% 
  filter(species == "Chinook") %>% 
  arrange(-catch)

DT::datatable(catch_chinook)
```

# Joins

load regions data table

```{r}
region_defs <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", method = "libcurl"),
                        stringsAsFactors = FALSE)


head(region_defs)
```

```{r}
region_defs <- region_defs %>% 
  select(code, mgmtArea)

head(region_defs)
```

```{r}
catch_joined <- left_join(catch_clean_mutated, region_defs, by = c("Region" = "code"))

head(catch_joined)

i <- which(is.na(catch_joined$mgmtArea))
nrow(i)
```


```{r}
dates_df <- data.frame(date = c("5/24/1930",
                                "5/25/1930",
                                "5/26/1930",
                                "5/27/1930",
                                "5/28/1930"),
                       stringsAsFactors = FALSE)

```


```{r}
dates_separate <- dates_df %>% 
  separate(date, into = c("month", "day", "year"), remove = F) %>% 
  unite(col = date_iso, year, month, day, sep = "-")

dates_separate

```

